---
title: "Exploring Land Value Disparities in New York City"
author: "Denis Folitse, Sandani Kumanayake, and Qi Chen"
date: "`r Sys.Date()`"
output:
  slidy_presentation: default
  beamer_presentation: default
  ioslides_presentation: default
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Introduction


## Research Questions

- Does the location of the property affect the Average land value in NYC?
- Which borough has the average highest land value?

## Data

- The Primary Land Use Tax Lot Output (PLUTO) is the data source used for this project.

- PLUTO is being provided by the Department of City Planning (DCP) on DCP’s website for informational purposes only.

- The original data set contains data of around 859877 properties in New York City.

- PLUTO includes 92 variables that provide detailed information on land use, building characteristics, and tax assessment values for each property in the city.


## Data Cleaning

- Original Data set had many missing values.
- Disproportionate stratified sampling was used to select a subset of data from the original PLUTO data set.
- The subset included 200 random properties from each borough, resulting in a new data set.
- Stratified sampling ensures that each borough is represented in the sample.


##

```{r, fig.width=10, fig.height=5, fig.align='center'}
library(DiagrammeR)

grViz("
  digraph flowchart {
    node [fontname = arial, shape = box, style = filled, fillcolor = lightblue, margin = 0.15, fontsize = 10]
    edge [fontname = arial, fontsize = 9]
    
    Data [label = 'Data Set\n859,877 properties\n92 variables']
    WithoutMissingValues [label = 'Data Set With Selected Variables\n(Without Missing Values)']
    Manhattan [label = 'Manhattan\n40098 properties']
    Bronx [label = 'Bronx\n83153 properties']
    Queens [label = 'Queens\n311881 properties']
    Brooklyn [label = 'Brooklyn\n265301 properties']
    StatenIsland [label = 'Staten Island\n115208 properties']
  
    
    Manhattan_1 [label = 'Manhattan\n200 properties']
    Bronx_1 [label = 'Bronx\n200 properties']
    Queens_1 [label = 'Queens\n200 properties']
    Brooklyn_1 [label = 'Brooklyn\n200 properties']
    StatenIsland_1 [label = 'Staten Island\n200 properties']
    SampleData [label = 'Sample Data Set\n1000 properties']
    
    Data -> WithoutMissingValues
    WithoutMissingValues -> Manhattan
    WithoutMissingValues -> Bronx
    WithoutMissingValues -> Queens
    WithoutMissingValues -> Brooklyn
    WithoutMissingValues -> StatenIsland
    
    Manhattan -> Manhattan_1
    Bronx -> Bronx_1
    Queens -> Queens_1
    Brooklyn -> Brooklyn_1
    StatenIsland -> StatenIsland_1
    
    Manhattan_1 -> SampleData
    Bronx_1 -> SampleData
    Queens_1 -> SampleData
    Brooklyn_1 -> SampleData
    StatenIsland_1 -> SampleData
  }
")


```


## Variables

- **Borough:** The borough in which the property is located.
- **Block:** The tax block in which the property is located. It contains a one to five-digit tax block number. Each tax block is unique within a borough.
- **lot:** The number of the tax lot. It contains a one to four-digit tax lot number. Each tax lot
  is unique within a tax block.
- **schooldist:** The school district in which the property lot is located.
- **policeprct:** The police precinct in which the property is located.
- **healthcenterdistrict:**	The health center district in which the property is located.
- **sanitdistrict:** The sanitation district that services the property.
- **ave_land:** Average price of the land (Dollars per Square Foot)
  

```{r, include=FALSE}
# Load packakages --------------------------------------------------------------
library(leaflet)
library(leaflet.extras)
library(sf)
library(tidyverse)
library(plotly)

# Read in Dataset and `kml` files ----------------------------------------------
dat = read.csv('dataset_group2.csv')
nyc_boundary = st_read('nyc_boundary.kml')
nyc_boundary = nyc_boundary %>% rename(borough = Name)
nyc_school = st_read('nyc_school.kml')
nyc_school = nyc_school %>% rename(schooldist = Name)

# Manipulate data --------------------------------------------------------------

# Change the name of boroughs in `dat`
dat_new = dat %>% 
  mutate(borough = str_replace(borough, 'BK', 'Brooklyn')) %>%
  mutate(borough = str_replace(borough, 'BX', 'Bronx')) %>%
  mutate(borough = str_replace(borough, 'MN', 'Manhattan')) %>%
  mutate(borough = str_replace(borough, 'QN', 'Queens')) %>%
  mutate(borough = str_replace(borough, 'SI', 'Staten Island'))
dat_new$schooldist = as.character(dat_new$schooldist)
dat_new$policeprct = as.character(dat_new$policeprct)

# Add three new columns `ave_borough`(average land value in corresponding borough), 
# `ave_school`(average land value in corresponding school district within each borough), 
# `ave_police`(average land value in corresponding police precinct district within 
# each borough) to `dat_new`.

dat_new = dat_new %>% group_by(borough) %>% mutate(ave_borough = mean(ave_land)) 
dat_new = dat_new %>% group_by(borough, schooldist) %>% mutate(ave_school = mean(ave_land))
dat_new = dat_new %>% group_by(borough, policeprct) %>% mutate(ave_police = mean(ave_land))


# For nyc_boundary: add column `ave_borough`
dat_borough = dat_new %>% group_by(borough) %>% summarise(ave_borough = mean(ave_land))
nyc_boundary = nyc_boundary %>% left_join(dat_borough, by = 'borough')

# For nyc_school: add column `ave_school`
dat_school = dat_new %>% group_by(borough, schooldist) %>% summarise(ave_school = mean(ave_land), .groups = 'drop')
nyc_school = nyc_school %>% left_join(dat_school, by = 'schooldist')

```

## Visualize the 5 boroughs in a map & Boxplot
- Rank of average land value: 
Manhattan > Brooklyn > Queens > Bronx > Staten Island
```{r}
# Plot leaflet map for 5 boroughs
#quantile(nyc_boundary$ave_borough)
bins = c(0, 4, 6.5, 7, 12, 142)
pal = colorBin('YlOrRd', domain = nyc_boundary$ave_borough, bins = bins)
labels <- sprintf(
  "<strong>Borough:</strong> %s<br/><strong>Average Land Value:</strong> %g",
  nyc_boundary$borough, nyc_boundary$ave_borough
) %>% 
  lapply(htmltools::HTML)
map = leaflet(nyc_boundary) %>%
  addTiles() %>%
  addPolygons(fillColor = ~pal(ave_borough),
              weight = 2,
              opacity = 1,
              color = 'white',
              dashArray = '3',
              fillOpacity = 0.7, 
              highlight = highlightOptions(weight = 5,
                                           color = "#666",
                                           dashArray = "",
                                           fillOpacity = 0.7,
                                           bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(style = list("font-weight" = "normal", 
                                                       padding = "3px 8px"),
                                          textsize = "15px",
                                          direction = "auto")) %>%
  #label = ~paste('Borough:', borough, '\nAverage Land Value:', round(ave_borough, 2))) %>%
  addLegend(pal = pal, values = ~ave_borough, opacity = 0.7, title = NULL, 
            position = "bottomright")
  

# Plot boxplot for `ave_land` by borough. 
box = dat_new %>% plot_ly(x = ~borough, y = ~ave_land, type = 'box', color = ~borough)
crosstalk::bscols(map, box)

```

## Add School Districts to the map
- Except Staten Island, all the other boroughs have more than one school districts
```{r}
# Plot leaflet map for school districts
labels <- sprintf(
  "<strong>Borough:</strong> %s<br/><strong>School District:</strong> %s<br/><strong>Average Land Value:</strong> %g",
  nyc_school$borough, nyc_school$schooldist, nyc_school$ave_school
) %>% 
  lapply(htmltools::HTML)
pal = colorFactor(c('#fed976', '#fc4e2a', '#bd0026', '#fd8d3c', '#ffffcc'), domain = nyc_school$borough)
leaflet(nyc_school) %>%
  addTiles() %>%
  addPolygons(fillColor = ~pal(borough),
              weight = 2,
              opacity = 1,
              color = 'white',
              dashArray = '3',
              fillOpacity = 0.7, 
              highlight = highlightOptions(weight = 5,
                                           color = "#666",
                                           dashArray = "",
                                           fillOpacity = 0.7,
                                           bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(style = list("font-weight" = "normal", 
                                                       padding = "3px 8px"),
                                          textsize = "15px",
                                          direction = "auto")) %>%
  #label = ~paste0('Borough: ', borough, 'School District: ', schooldist,  ' Average Land Value: ', round(ave_school, 2))) %>%
  addLegend(pal = pal, values = ~borough, opacity = 0.7, 
            title = NULL, position = "bottomright")
#range(nyc_school$ave_school)

```

## Bar charts of four categorical variables VS. boroughs

```{r}
# Plot bar chart for number of school districts by borough
borough = c('Bronx', 'Brooklyn', 'Manhattan', 'Queens', 'Staten Island')
school_bar = numeric(5)
j = 1
for (i in borough) {
  school_bar[j] = nrow(dat_new %>% group_by(borough, schooldist) %>% 
                         summarise(n = n(), .groups = 'drop') %>% filter(borough == i))
  j = j + 1
}
#school_bar
bar_shcool = plot_ly(x = borough, y = school_bar, legendgroup = ~borough, 
                      color = borough, type = 'bar') %>% 
    add_text(
      text = school_bar,
      textfont = list(color = 'red'),
      textposition = 'bottom center',
      showlegend = FALSE,
      cliponaxis = FALSE,
      hoverinfo = 'none'
    )

# Plot bar char for number of police precinct district by borough
police_bar = numeric(5)
j = 1
for (i in borough) {
  police_bar[j] = nrow(dat_new %>% group_by(borough, policeprct) %>% 
                         summarise(n = n(), .groups = 'drop') %>% filter(borough == i))
  j = j + 1
}
#police_bar
bar_police = plot_ly(x = borough, y = police_bar, legendgroup = ~borough, 
                      color = borough, type = 'bar') %>% 
    add_text(
      text = police_bar,
      textfont = list(color = 'red'),
      textposition = 'bottom center',
      showlegend = FALSE,
      cliponaxis = FALSE,
      hoverinfo = 'none'
    )

# Plot bar char for number of health center districts by boroughs
health_bar = numeric(5)
j = 1
for (i in borough) {
  health_bar[j] = nrow(dat_new %>% group_by(borough, healthcenterdistrict) %>% 
                          summarise(n = n(), .groups = 'drop') %>% filter(borough == i))
  j = j + 1
}
#health_bar
bar_health = plot_ly(x = borough, y = health_bar, legendgroup = ~borough, 
                       color = borough, type = 'bar') %>% 
    add_text(
      text = health_bar,
      textfont = list(color = 'red'),
      textposition = 'bottom center',
      showlegend = FALSE,
      cliponaxis = FALSE,
      hoverinfo = 'none'
    )

# Plot bar chart for number of sanitation districts by boroughs 
sani_bar = numeric(5)
j = 1
for (i in borough) {
  sani_bar[j] = nrow(dat_new %>% group_by(borough, sanitdistrict) %>% 
                         summarise(n = n(), .groups = 'drop') %>% filter(borough == i))
  j = j + 1
}
#sani_bar
bar_sani = plot_ly(x = borough, y = sani_bar, legendgroup = ~borough, 
                      color = borough, type = 'bar') %>% 
    add_text(
      text = sani_bar,
      textfont = list(color = 'red'),
      textposition = 'bottom center',
      showlegend = FALSE,
      cliponaxis = FALSE,
      hoverinfo = 'none'
    )

# Organize the above four plots in one page
subplot(style(bar_shcool, showlegend = F), style(bar_police, showlegend = F),  
        style(bar_health, showlegend = F), bar_sani, nrows = 2, margin = 0.08) %>%
  layout(annotations = list( 
    list(text = "Number of School Districts VS. Boroughs", x = 0, y = 1.03, 
         showarrow = F, xref='paper', yref='paper', font = list(family = "Times New Roman", size = 13)), 
    list(text = "Number of Police Precinct Districts VS. Boroughs", x= 1, y = 1.03, 
         showarrow = F, xref='paper', yref='paper', font = list(family = "Times New Roman", size = 13)),
    list(text = 'Number of Health Center Districts VS. Boroughs', x = 0, y = 0.46, showarrow = F,  
         xref='paper', yref='paper', font = list(family = "Times New Roman", size = 13)),
    list(text = 'Number of Sanitation Districts VS. Boroughs', x = 1, y = 0.46, showarrow = F,  
         xref='paper', yref='paper', font = list(family = "Times New Roman", size = 13)))
  ) 
```



## Shiny app

```{r}
setwd("/Users/sandanikumanayake/Desktop/STAT 697/GP 02/")
shiny::shinyAppFile('app_v3.R')
```


## Conclusions

- As we observed, the value (price) of land in New York is influenced significantly by it's location with Manhattan Borough having averagely the most priced land.

- In Manhattan, school district 2, is generally associated with high land value as compared to the others.

- Similarly, 14th to 18th police precincts;  health centered districts 13 and 15, sanitation district 5 are associated with high values of lands in Manhattan.  

