---
title: "Relation between College Costs and Expected Earnings"
author: "Edward Harvey, Yifan Li, Sandani Kumanayake"
date: "`r Sys.Date()`"
output:
  slidy_presentation: default
  beamer_presentation: default
  ioslides_presentation: default
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```




# Introduction

## Question

- Do high college costs mean high pay-off in future earnings?

- What about other factors? 
  + percentage of STEM majors
  + diversity

## Data

Accessed via Kaggle.com. Four files from two sources:

-	Diversity by college/university for 2014, from the Chronicle of Higher Education (4575 unique values)

-	Tuition and fees by college/university for 2018-2019, from the Chronicle of Higher Education (2 datasets) (2938 and 3664 unique values)

-	Salary potential data comes, from payscale.com (934 unique values)

## Variables

- Costs 
  + in_state_total
  + out_of_state_total
- Expected Earnings
  + early_career_pay
  + mid_career_pay
- Others
  + stem_percent
  + pct_women
  + pct_minority

## Weaknesses

- Different time periods.

- Not whole population.

- Many NA values.

## Cleaning

1. Starting with diversity data
    +	Selected university name, state, and total enrollment variables
    +	Pivoted wider by name and state to get counts for different student demographics
2. Joined salary potential data with newly pivoted diversity data based on university name and state 
3. Jointed tuition cost data to new data frame
4. Cleaned second tuition cost data and joined to new data frame
    +	Pivoted wider to get cost breakdown by income level 
    +	Joined with new data frame by name and state
    +	Ultimately did not use this data in the analysis
5. Removed NAs as appropriate for individual plots
6. Calculated percentages, geographic regions, and other minor variable changes as necessary for individual plots


```{r, include=FALSE}
library(tidyverse)
library(usdata)
library(ggplot2)
library(plotly)
library(forcats)
library(shiny)
library(ggpubr)
```


```{r cleanup, echo=TRUE}
setwd("/Users/sandanikumanayake/Desktop/STAT 697/HW 05/Final/")

#read in diversity dataset
diversity <- read.csv2("diversity_school.csv", sep=",")

#separate total enrollment before pivot, take distinct entries by name and state
total_enrollment <- diversity %>% select(name, state, total_enrollment) %>% 
  distinct()

#drop total enrollment from diversity
diversity <- diversity %>% select(-total_enrollment)

#convert category and state to factor (not sure if this makes a difference?)
diversity$category <- as.factor(diversity$category)
diversity$state <- as.factor(diversity$state)


#pivot demographic data (saved as new df)
diversity_pivot <- diversity %>% 
  filter(!is.na(name)) %>% 
  pivot_wider(id_cols=c(name, state), names_from=category, values_from = enrollment)

#here I rejoin the total enrollment and diversity_pivot is complete

diversity_pivot <- left_join(diversity_pivot, total_enrollment, by=c("name", "state"))


#load in the salary potential data
salary_potential <- read.csv2("salary_potential.csv", sep=",")
salary_potential <- salary_potential %>% rename("state" = "state_name")

#join diversity and salary potential, save as new df
df <- left_join(diversity_pivot, salary_potential, by=c("name", "state"))

#join df with tuition_cost
tuition_cost <- read.csv2("tuition_cost.csv", sep=",")
df <- left_join(df, tuition_cost, by=c("name", "state"))

#drop state code (redundant) and rank(?)
df <- df %>% select(-c("state_code","rank"))


#load tuition income dataset
tuition_income <- read.csv2("tuition_income.csv", sep=",")

#select 2018 only, the most recent data (other datasets do not have multiple years)
tuition_income <- tuition_income %>% 
  filter(year==2018) %>% 
  select(name, state, total_price, campus) %>% 
  distinct()

#convert to factor (is this necessary?)
tuition_income$campus[which(tuition_income$name=="Laurel Technical Institute" & tuition_income$total_price==16168)] <- 'On Campus'

tuition_income$campus <- as.factor(tuition_income$campus)
  
#pivot wider
tuition_income_pivot <- tuition_income %>% pivot_wider(names_from = campus, values_from = total_price)

#change null values to NA for On Campus
tuition_income_pivot$`On Campus`[sapply(tuition_income_pivot$`On Campus`, is.null)] <- NA

#convert state codes to state
tuition_income_pivot$state <- abbr2state(tuition_income_pivot$state)


#join 
df <- left_join(df, tuition_income_pivot, by=c("name", "state"))


colnames(tuition_income_pivot[,3:4]) <- c("on_campus","off_campus")

#save as RDS file
write.csv2(df, file="df.csv")

saveRDS(df, file="df.rds")

```


# Plots

## Costs

```{r, fig.width=10, fig.height=5, fig.align='center'}
fig <- df %>%
  filter(type != "For Profit") %>%
  plot_ly(type = 'violin') 

fig <- fig %>%
  add_trace(
    x = ~type,
    y = ~in_state_total,
    legendgroup = 'In State',
    scalegroup = 'In State',
    name = 'In State',
    side = 'negative',
    box = list(
      visible = T
      ),
    meanline = list(
      visible = T
      ),
    color = I("blue")
    ) 

fig <- fig %>%
  add_trace(
    x = ~type,
    y = ~out_of_state_total,
    legendgroup = 'Out of State',
    scalegroup = 'Out of State',
    name = 'Out of State',
    side = 'positive',
    box = list(
      visible = T
      ),
    meanline = list(
      visible = T
      ),
    color = I("green"
    ) )

fig %>%
  layout(
    xaxis = list(
      title = ""
      ),
    yaxis = list(
      title = "Total Cost",
      zeroline = F
      ),
    # violingap = 0,
    # violingroupgap = 0,
    violinmode = 'overlay'
    )
```

## Expected Earnings

```{r}
df2 = df
# Convert early_career_pay column to numeric
df2$early_career_pay <- as.numeric(df2$early_career_pay)

# Calculate median early_career_pay by type (Used median because distributions are not symmetric)
median_values <- aggregate(early_career_pay ~ type, data = na.omit(df2), FUN = median)

# Get the total number of colleges
total_colleges <- nrow(na.omit(df2))

# Define colors for median lines
median_colors <- c("darkgreen", "darkorange")

```

```{r, fig.width=10, fig.height=5, fig.align='center'}
# Plot histogram of early_career_pay by type without NA values
ggplot(data = na.omit(df2), aes(x = early_career_pay, color = type, fill = type)) +
  geom_histogram(aes(y = ..count..), binwidth = 5000, color = "black", alpha = 0.5, position="identity") +
  geom_vline(data = median_values, aes(xintercept = early_career_pay, color = type),
             linetype = "dashed", size = 1) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_manual(values = median_colors) +
  theme_minimal() +
  labs(x = "Early Career Pay ($)", y = "Count", fill = "Type") +
  ggtitle("Distribution of Early Career Pay by Type") +
  theme(plot.title = element_text(size = 20, face = "bold"),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))+
  guides(fill = guide_legend(order = 1), color = guide_legend(order = 2, title = "Median"))
```

## Scatter Plots

```{r}
#named vectors for explanatory and response variables
explanatory_choices <-
  c(
    "Total Cost" = "out_of_state_total",
    "% STEM" = "stem_percent",
    "% Women" = "pct_women",
    "% Minority" = "pct_minority"
  )

response_choices <-
  c(
    "Early-Career Earnings" = "early_career_pay",
    "Mid-Career Earnings" = "mid_career_pay"
  )



#load the data
df <- readRDS("df.rds")

#convert deomgraphic data to percentages
df$pct_minority <- df$`Total Minority`/df$total_enrollment
df$pct_women <- df$Women/df$total_enrollment

#add variable for region
df$region <- setNames(state.region, state.name)[df$state]
```


```{r}
# Run the application 
shinyAppFile("app_v3.R")
```

## Conclusion

- Some of the more expensive private universities show higher career earnings, but for many this is not the case

- Public universities show a more positive and linear relationship between costs and future earnings

- Strong positive relationship between % STEM and future earnings

- Universities with highest future earnings have a roughly balanced gender mix

- Male-dominated schools tend to be technical universities with higher future earnings compared to female-dominated schools

